---
title: "THE POPULARITY OF BOOKS ON GOODREADS"
subtitle: "DATA 606 Final Project"
author: "Stephanie Chiang"
date: "FALL 2024"
output:
  pdf_document: default
  html_document: default
urlcolor: blue
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Abstract

This is a statistical analysis of some of the factors that contribute to the popularity of books on [Goodreads](goodreads.com). This observational dataset posted on Kaggle was originally sourced from Goodreads' Top 100 lists of the most popular books for each year from 1980 to 2023. In particular, the focus will be to answer the question:

Are serials, or books that are part of multi-volume series, associated with higher ratings, more reviews or a larger current or potential readership pool than standalone books? In other words, is there a relationship between the independent variables (standalone books vs serials, and first in series vs sequel) and the dependent variables (mean user ratings, number of ratings, and readership counts)?

Using statistics, visualizations and regression modeling, the results indicate that serials are not higher average ratings, with sequels (or prequels) receiving more ratings and higher average ratings than a first published installment. However, there are more current and potential Goodreads users interested in reading standalone books than books in a series. This analysis provides potentially valuable insights for publishers, authors, and marketers for data-driven decision-making.


### Data Preparation

The explanatory variables are created from text fields in the raw data. Each of the 4399 rows is given a value in a new categorical column with two levels (TRUE or FALSE) called `serial` based on whether there are non-empty strings under `series_title` and `series_release_number`.

Then, in a second data frame for only those books with `serial` set to TRUE, each observation is marked in a new `first_book` column with a TRUE if it is the first book published in its series or FALSE for sequels and prequels. The determination here is that since prequels are published *after* the initial volume, they should not be considered the first in a series.

The response variables are numerical: `rating_score`, `num_ratings`, `current_readers`, `want_to_read`.

```{r import, message=FALSE, warning=FALSE}
raw_books <- read.csv(file = "goodreads_top100.csv")

# select relevant columns
books <- raw_books |>
  select("isbn",
         "title",
         "series_title",
         "series_release_number",
         "rating_score",
         "num_ratings",
         "current_readers",
         "want_to_read")
```

```{r clean-books-df}
# convert blank strings to NAs in text columns
books <- books |>
  mutate(isbn = na_if(isbn, "")) |>
  mutate(series_title = na_if(series_title, "")) |>
  mutate(series_release_number = na_if(series_release_number, ""))

# remove duplicate ISBN numbers
books <- books |>
  distinct(isbn, .keep_all = TRUE)

# add column to indicate if the book in series
books <- books |>
  mutate(serial = !is.na(books$series_title) & !is.na(books$series_release_number))

head(books)
```

```{r serial-df}
# create a 2nd table for any series-only analysis
series <- filter(books, serial == TRUE)

# add column for if it is the first release of its series
series <- series |>
  mutate(first_book = ifelse(grepl("^1(?!\\d)", series$series_release_number, perl = TRUE),
                             TRUE,
                             FALSE)) |>
  subset(select = -c(serial))

head(series)
```

### Summary Statistics & Data Visualizations

*Mean Ratings: Series vs Standalone Books*

Comparing the average ratings shows only a slight a preference by Goodreads users for serials over standalone books.

```{r series-ratings, message=FALSE, warning=FALSE}
rating_summary <- books |>
  group_by(serial) |>
  reframe(
    count = n(),
    mean = mean(rating_score),
    sd = sd(rating_score),
    median = median(rating_score),
    min = min(rating_score),
    max = max(rating_score),
  )

rating_summary
```

```{r series-box}
ggplot(books, aes(x = serial, y = rating_score)) +
  geom_boxplot() +
  scale_x_discrete(labels = c("FALSE" = "Standalone book", "TRUE" = "Part of a series")) +
  labs(y = "Goodreads average rating", x= "")
```

Since each book is an independent observation and the sample sizes for each group are comfortably large, the conditions for inference are satisfied; a hypothesis test for the difference of the two means can evaluate the relationship.

The null hypothesis H0: There is no relationship between being part of a series and average rating.
The alternative hypothesis H1: The average ratings are significantly different for serials.

Below, the difference in means is calculated in the order of TRUE - FALSE != 0.
The test is then simulated on the null distribution and plotted.

```{r series-hyp, message=FALSE, warning=FALSE}
library(infer)
set.seed(99)

series_obs_diff <- books |>
  specify(rating_score ~ serial) |>
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

series_null_dist <- books |>
  specify(rating_score ~ serial) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

ggplot(data = series_null_dist, aes(x = stat)) +
  geom_histogram()
```

```{r series-pval, message=FALSE, warning=FALSE}
series_null_dist |>
  get_p_value(obs_stat = series_obs_diff, direction = "two_sided")

series_diff_ci <- series_null_dist |> get_ci(level = 0.95)

series_diff_ci
```

At a confidence interval of 95%, the difference in mean ratings between series and standalone books should fall between -0.014 to 0.015. Since this contains 0, we can fail to reject the null hypothesis. There is no significant difference in average rating for serials.


*Average Ratings: Firsts vs Sequels*

Within series, there is a somewhat more noticeable bump in ratings for sequels over the first book.

```{r first-ratings}
rating_summary_sequels <- series |>
  group_by(first_book) |>
  reframe(
    count = n(),
    mean = mean(rating_score),
    sd = sd(rating_score),
    median = median(rating_score),
    min = min(rating_score),
    max = max(rating_score),
  )

rating_summary_sequels
```

```{r first-box}
ggplot(series, aes(x = first_book, y = rating_score)) +
  geom_boxplot() +
  scale_x_discrete(labels = c("FALSE" = "Sequel or prequel", "TRUE" = "First of its series")) +
  labs(y = "Goodreads average rating", x= "")
```

The hypothesis test for serials alone is as follows: 

The null hypothesis H0: There is no relationship between being a sequel and average rating.
The alternative hypothesis H1: The average ratings are significantly different for sequels than first books.

```{r first-hyp, message=FALSE, warning=FALSE}
set.seed(99)

first_obs_diff <- series |>
  specify(rating_score ~ first_book) |>
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

first_null_dist <- series |>
  specify(rating_score ~ first_book) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

ggplot(data = first_null_dist, aes(x = stat)) +
  geom_histogram()
```

```{r first-pval, message=FALSE, warning=FALSE}
first_null_dist |>
  get_p_value(obs_stat = first_obs_diff, direction = "two_sided")

first_diff_ci <- first_null_dist |> get_ci(level = 0.95)

first_diff_ci
```

At a confidence interval of 95%, the difference in mean ratings between first books and sequels/prequels should fall between -0.02 to 0.02. Since this contains 0, we can fail to reject the null hypothesis. There is no significant difference in average rating for sequels.


*Readership*

Here is a comparison of users who marked themselves as current readers of a title vs interested/potential readers. As visualized below, the mean number of users who `want_to_read` a standalone book is much higher than for series, even though the ratings tend to be lower.
 
```{r readership-count}
books_readership <- books |>
  pivot_longer(cols = c("current_readers", "want_to_read"),
               names_to = "reader_type",
               values_to = "readership")

readership_summary <- books_readership |>
  group_by(serial, reader_type) |>
  summarize(mean_readership = mean(readership, na.rm = TRUE))

readership_summary
```

```{r readership-bar}
ggplot(readership_summary, aes(fill = reader_type, x = serial, y = mean_readership)) + 
  geom_bar(position = "dodge", stat = "identity") +
  scale_y_sqrt() +
  labs(y = "Number of users", x= "") +
  scale_x_discrete(labels = c("FALSE" = "Standalone book", "TRUE" = "Part of a series")) +
  scale_fill_discrete(name = "Reader type", labels = c("Current Reader", "Want to read"))
```

For regression models, include the regression output and interpret the R-squared value.



### Conclusion
Why is this analysis important?
Limitations of the analysis?
- Conclusion includes a clear answer to the statistical question that is consistent with the data analysis and the method of data collection.

