---
title: "DATA 606 Data Project Proposal"
author: "Stephanie Chiang"
output: pdf_document
urlcolor: blue
date: "2024-11-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r imports, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

### Data Preparation

```{r import, message=FALSE, warning=FALSE}
raw_books <- read.csv(file = "goodreads_top100.csv")

# select relevant columns
books <- raw_books |>
  select("isbn",
         "title",
         "series_title",
         "series_release_number",
         "rating_score",
         "num_ratings",
         "current_readers",
         "want_to_read")
```

```{r clean-books-df}
# convert blank strings to NAs in text columns
books <- books |>
  mutate(isbn = na_if(isbn, "")) |>
  mutate(series_title = na_if(series_title, "")) |>
  mutate(series_release_number = na_if(series_release_number, ""))

# remove duplicate isbn numbers
books <- books |>
  distinct(isbn, .keep_all = TRUE)

# add column to indicate if in series
books <- books |>
  mutate(serial = !is.na(books$series_title) & !is.na(books$series_release_number))

head(books)
```

```{r serial-df}
# create a 2nd table for any serials-only analysis
serial_books <- filter(books, serial == TRUE)

# add column for if it is first in a series
# may need further cleanup, depends if a prequel counts as first?
serial_books <- serial_books |>
  mutate(first_book = ifelse(grepl("^[01]", serial_books$series_release_number),
                             TRUE,
                             FALSE)) |>
  subset(select = -c(serial))

head(serial_books)
```


### Research question 

Are serials (books that are part of multi-volume series) more popular (either by readership, interest or rating) than standalone books? In other words, is there a relationship between being part of a series and popularity for books? 


### Cases 

Each case is a book published between 1980-2023 and ranked in the top 100 for its year, based on Goodreads ratings. There are 4399 cases.


### Data collection 

This dataset posted on Kaggle was "collected through web scraping techniques" from Goodreads.com


### Type of study 

This is an observational study.


### Data Source 

[Link to Top Goodreads Books Collection](https://www.kaggle.com/datasets/cristaliss/ultimate-book-collection-top-100-books-up-to-2023)


### Describe your variables?

The response variables are numerical: rating_score, num_ratings, current_readers, want_to_read.
The explanatory variables are text fields, which will be converted into boolean / categorical: series_title (exists yes/no), series_release_number (first in series or not)


### Relevant summary statistics 

```{r summary, message=FALSE, warning=FALSE}
# summary statistics for rating_score, serials vs standalone
rating_summary <- books |>
  group_by(serial) |>
  reframe(
    count = n(),
    mean = mean(rating_score),
    sd = sd(rating_score),
    median = median(rating_score),
    min = min(rating_score),
    max = max(rating_score),
  )

rating_summary

# summary statistics for rating_score, first in a series vs sequel
rating_summary_sequels <- serial_books |>
  group_by(first_book) |>
  reframe(
    count = n(),
    mean = mean(rating_score),
    sd = sd(rating_score),
    median = median(rating_score),
    min = min(rating_score),
    max = max(rating_score),
  )

rating_summary_sequels

# comparing mean of the numbers of current readers vs interested/potential readers
books_readership <- books |>
  pivot_longer(cols = c("current_readers", "want_to_read"),
               names_to = "reader_type",
               values_to = "readership")

readership_summary <- books_readership |>
  group_by(serial, reader_type) |>
  summarize(mean_readership = mean(readership, na.rm = TRUE))

readership_summary
```

```{r box-plots}
ggplot(books, aes(x = serial, y = rating_score)) +
  geom_boxplot() +
  labs(x = "Is the book part of a series?", y = "Goodreads rating")

ggplot(serial_books, aes(x = first_book, y = rating_score)) +
  geom_boxplot() +
  labs(x = "Is the book the first in its series?", y = "Goodreads rating")

ggplot(readership_summary, aes(fill = reader_type, x = serial, y = mean_readership)) + 
  geom_bar(position = "dodge", stat = "identity") +
  scale_y_sqrt() +
  labs(x = "Is the book part of a series?", y = "Readership")
```

